Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array: inclusion of pulsar terms

### 2023-11-13 manuscript/manuscript_2/paper.tex
assorted manuscript updates

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[fleqn,usenatbib,useAMS]{mnras}

@@

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
	\title[Kalman PTA]{Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms}

@@

% Abstract of the paper
\begin{abstract}	
	The detection of continuous nanohertz gravitational waves from individual supermassive black hole binaries can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival. Such an approach complements traditional analysis methods by enabling self-consistent inference of model parameters and calculation of the Bayesian evidence, without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise. Previous analyses have worked exclusively with the `Earth term', leading to biases in the parameter estimation. Here it is shown how to extend previous analyses to include the so-called `pulsar terms'. It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy relative to the Earth terms approach by X and the sensitivity by Y. 
\end{abstract}

@@

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
Nanohertz gravitational waves (GWs) can be produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}. A nHz GW influences the times of arrival (TOAs) of radio pulses from pulsars measured at the Earth. A pulsar timing array \citep[PTA][]{Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars. In this way it is possible to detect GWs at frequencies that are prohibitively low for terrestrial interferometers. \newline

 PTAs search for two types of SMBHB GW signals. The first is a stochastic GW background due to the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. Consilient evidence for the GW background has been presented by multiple separate PTAs \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X}, by measuring the cross-correlation between pairs of pulsars in the pulsar timing residuals as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}. The second is resolved GWs from sufficiently close and massive individual SMBHBs \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. Individual SMBHBs are sources of continuous GWs. This results in persistent, quasi-monochromatic modulations in the pulsar timing residuals which can be modelled. Therefore, rather than cross-correlating pulsar pairs, individual SMBHBs are detected by either frequentist matched filtering \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or Bayesian inference \citep{Ellis2016,Arzoumanian2020A}. Observational efforts to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, although some inconclusive evidence at low significance has been presented \citep{2023arXiv230616226A}. \newline

 Irregularities in the intrinsic rotation of the pulsar result in random, unmodelled, red-spectrum TOA fluctuations known as `timing noise'. Timing noise limits the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}. Additional white noise sources such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021} must also be considered. \cite{K1} (K23 hereafter) presented an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations. Specifically, K23 formulate PTA analysis as a state-space problem and use a Kalman filter \citep{Kalman1} to track the pulsars' intrinsic rotational state in conjunction with a Bayesian nested sampler \citep{Skilling, Ashton2022} for parameter estimation and model selection. \textbf{VALIDATION TESTS ON SYNTHETIC DATA... }newline

 It is common in standard PTA analysis to drop the so called ``pulsar term" from the modelling \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}. Unlike the "Earth term", the pulsar term is not phase coherent and cannot be summed across the array to increase the total signal-to-noise. Instead it is considered as a source of self noise. However, dropping the pulsar terms results in a reduced detection probability ($\sim 5 \%$) bias in the parameter estimates \citep{Zhupulsarterms,Chen2022}. The method presented in K23 also drops the pulsar term. In this paper we remove this limitation and present the state-space approach to PTA data analysis including the pulsar term. \newline

 This paper is organised as follows. In Section \ref{sec:model} we briefly review the state-space formulation of PTA data analysis introduced in \cite{K1}. We do not review the methods for state tracking, parameter inference and model selection and refer the reader to K23 for details. In Section \ref{sec:pulsar_term} we introduce how to include the pulsar term in the state-space model. In Section \ref{sec:rep_example1} we test the updated method on the synthetic data for a single representative example GW system, and compare the results to the original Earth-terms model. In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space. Discussion on future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively. Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$. \newline

% Could include the below block if we need more fodder

%This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}. In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators. However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}.

\section{State-Space Formulation}\label{sec:model}
In this section we briefly review the state-space formulation of PTA analysis presented in K23. \newline

There are $N$ pulsars in the array. The index $n$ labels the $n$-th pulsar, with $1\leq n\leq N$. The intrinsic spin frequency of the $n$-th pulsar as measured in the local, freely-falling rest frame of the pulsar's centre of mass, $f_{\rm p}^{(n)}(t)$, evolves according to a stochastic differential equation. The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation. In Section \ref{sec:psr_frequency} we review the phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:psr_measured} we review the measurement equation which relates $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:ss_params} we summarise the parameters of the model.

\subsection{Spin evolution} \label{sec:psr_frequency}
In the model $f_{\rm p}^{(n)}(t)$ evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ ,
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:
\begin{align}
	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
\end{align}
In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes noise amplitude and the gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}. PTAs are typically composed of millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and can be approximated accurately by 
\begin{equation}
	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} 
where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. \newline

Equation \ref{eq:frequency_evolution} is inherently an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}. Such a phenomenological model is necessary since a predictive, first-principles theory of timing noise does not currently exist c.f. the multiple theorized mechanisms of timing noise, see the discussion in Section 1 of K23. Similar approaches, analogous to Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} have been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}. However, Equation \eqref{eq:frequency_evolution} does contain certain modelling omissions and idealisations which much be recognized \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}. These include: exclusion of phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}, the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}, the suitability of the white noise driver $\xi(t)$ for millisecond pulsars in PTAs and the formal interpretation of $d^2 f_{\rm p} / dt^2$.

%
%However, as discussed in K23, First, Equation \ref{eq:frequency_evolution} excludes particular phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}. Second, the white spectrum assumed for $\xi(t)$ may or may not be suitable for millisecond pulsars in PTAs. It is challenging observationally to infer the spectrum of $\xi(t)$ from the observed spectrum of the phase residuals, because the inference is conditional on the (unknown) dynamical model governing $df_{\rm p}/dt$. For small-amplitude fluctuations sampled relatively often, as in millisecond pulsars in PTAs, it is likely that $\xi(t)$ is white to a good approximation over the inter-TOA intervals and generates red phase residuals as observed, but caution is warranted nevertheless. White noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Equation \eqref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments. Third, $\xi(t)$ is not differentiable, which makes the formal interpretation of $d^2 f_{\rm p} / dt^2$ ambiguous, even though $d^2 f_{\rm p} / dt^2$ is not used in the PTA analysis proposed in this paper. Finally, the Brownian increment $dB(t)=\xi(t)dt$ does not include non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}, which have not been ruled out by pulsar timing experiments to date. The above three idealizations are supplemented by other, physical approximations noted above, e.g.\ neglecting $n_{\rm em}$ in Equation \eqref{eq:spinevol} and differential rotation between the crust and superfluid in Equation \eqref{eq:frequency_evolution}.

%This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}. In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators. However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}. 
%
%

%
%Equations \eqref{eq:spinevol}--\eqref{eq:xieqn} are not derived from first principles by applying a microphysical theory. As a first pass, they also exclude certain phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}. 
%
%
%
%However, Equations \eqref{eq:spinevol}--\eqref{eq:xieqn}  involve significant idealizations, which must be recognized at the outset \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}. First, the white noise driver $\xi(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable, which makes the formal interpretation of $d^2 f_{\rm p} / dt^2$ ambiguous, even though $d^2 f_{\rm p} / dt^2$ is not used in the PTA analysis proposed in this paper. Second, the white spectrum assumed for $\xi(t)$ may or may not be suitable for millisecond pulsars in PTAs. It is challenging observationally to infer the spectrum of $\xi(t)$ from the observed spectrum of the phase residuals, because the inference is conditional on the (unknown) dynamical model governing $df_{\rm p}/dt$. For small-amplitude fluctuations sampled relatively often, as in millisecond pulsars in PTAs, it is likely that $\xi(t)$ is white to a good approximation over the inter-TOA intervals and generates red phase residuals as observed, but caution is warranted nevertheless. Third, the Brownian increment $dB(t)=\xi(t)dt$ does not include non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}, which have not been ruled out by pulsar timing experiments to date. The above three idealizations are supplemented by other, physical approximations noted above, e.g.\ neglecting $n_{\rm em}$ in Equation \eqref{eq:spinevol} and differential rotation between the crust and superfluid in Equation \eqref{eq:frequency_evolution}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
%\documentclass[fleqn,usenatbib,useAMS,onecolumn]{mnras}
\documentclass[fleqn,usenatbib,useAMS]{mnras}

@@

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
	\title[Kalman PTA]{Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array: inclusion of pulsar terms}

@@

% Abstract of the paper
\begin{abstract}

	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries.

	 The gravitational wave signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively.

	 	 The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival.

We show how to extend previous state-space analyses to include the pulsar terms.

It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy relative to the Earth terms approach by X and the sensitivity by Y

\end{abstract}

@@

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
Nanohertz gravitational waves (GWs) are produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}. A nHz GW influences the time of arrival (TOA) at the Earth of radio pulses from pulsars. A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars. In this way it is possible to detect GWs at frequencies that are prohibitively low for terrestrial interferometers. Consilient evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. SMBHBs which are sufficiently close and massive may be individually resolvable, though conclusive evidence is currently lacking \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}. \newline

The modulation in the pulsar TOAs due to the presence of a GW has two contributions: an ``Earth term" and a ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases. Consequently the pulsar terms are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}. However, dropping the pulsar terms leads to biases in the inferred parameters and reduced detection probabilities \citep{Zhupulsarterms,Chen2022,KimpsonPTA}. \newline

\cite{KimpsonPTA} (K24 hereafter) introduced a novel, state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, that self-consistently tracks the intrinsic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations. The method described in K24 follows the example of standard  PTA analyses and also drops the pulsar term. In this paper we undertake two key tasks. Firstly we extend the methods presented in K24 to include the pulsar term. We show how to reparametrise the model in terms of an additional pulsars-specific phase term. We demonstrate how the static GW parameters and the additional pulsar phase term can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}. Secondly we quantify how the inclusion of the pulsar term affects the accuracy with which the static GW parameters can be estimated, and the minimal detectable GW strain.

This paper is organised as follows. In Section \ref{sec:2} we briefly review the state-space formulation of PTA data analysis introduced in \cite{KimpsonPTA}.

Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$. \newline

%In Section 3 we introduce how to include the pulsar term in the state-space model. In Section 4 we test the updated method on the synthetic data for a single representative example GW system, and compare the results to the original Earth-terms model. In Section 9 we extend the tests to cover an astrophysically relevant parameter space. Discussion on future extensions and conclusions are made in Sections 11 and 12 respectively. Throughout the paper we adopt the natural units, with 𝑐 = 𝐺 = ħ = 1, and metric signature (−, +, +, +).

%
%
%Intrinsic pulsar timing noise -- i.e.  random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}. This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Myers2021MNRAS.502.3113M,Meyers2021,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}. In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators. However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}. In modern Bayesian PTA searches, the power spectral density of the red intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the noise induced by a stochastic GW background (whose spectrum is also red). In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

%In this work we present an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations. The new approach differs from existing approaches in one key respect: it infers the GW parameters conditional on the unique, time-ordered realization of the noisy TOAs observed, instead of fitting for the ensemble-averaged statistics of the TOA noise process, e.g., the amplitude and exponent of its power spectral density. Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations summarized by the power spectral density, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization. The new and existing approaches are therefore complementary. In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}. We combine the Kalman tracking of the pulsars' intrinsic rotational states with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the static GW parameters and calculate the marginal likelihood (i.e. the model evidence) for model selection. \newline

%The paper is organised as follows. In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW. In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence. In Section \ref{sec:testing} we describe how we create synthetic validation data with which to test the method. We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative GW source. In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant domain of SMBHB source parameters. In Section \ref{sec:bias_and_identifiability} we discuss identifiability of and bias in the parameter estimates. Conclusions are drawn in Section \ref{sec:discussion}.

\section{State-Space Formulation}\label{sec:2}
In this section we briefly review the state-space formulation of PTA analysis presented in K24. \newline

There are $N$ pulsars in the array. The index $n$ labels the $n$-th pulsar, with $1\leq n\leq N$. The intrinsic spin frequency of the $n$-th pulsar as measured in the local, freely-falling rest frame of the pulsar's centre of mass, $f_{\rm p}^{(n)}(t)$, evolves according to a stochastic differential equation. The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation which quantifies the influence of the GW. In Section \ref{sec:psr_frequency} we outline the phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:psr_measured} we outline the measurement equation which relates $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:ss_params} we summarise the parameters of the model. The model is formulated in terms of pulse frequency time series as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}. It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data. We defer this subtle generalization to future work.

\subsection{Spin evolution} \label{sec:psr_frequency}
The intrinsic pulsar spin frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ . 
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:
\begin{align}
	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
\end{align}
In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes noise amplitude and the gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}. PTAs are typically composed of millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and can be approximated accurately by 
\begin{equation}
	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} 
where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. \newline

Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} are inherently an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}. Such a phenomenological model is necessary as a predictive, first-principles theory of timing noise does not currently exist c.f. the multiple theorized mechanisms of timing noise (see discussion in Section 1 of \cite{KimpsonPTA}). Similar approaches, analogous to Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} have been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}. However, Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} do contain certain modelling omissions and idealisations which much be recognized \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}. These include the exclusion of phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; the suitability of the white noise driver $\xi(t)$ for millisecond pulsars in PTAs; and the formal interpretation of $d^2 f_{\rm p} / dt^2$ as $\xi(t)$ in Equation \eqref{eq:frequency_evolution} is not differentiable.

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the intrinsic pulsar spin frequency for the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a a non-linear measurement equation:
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}(t)\ ,
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ ,	\label{eq:vareps}
\end{align}
where $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig}
\end{align}
where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\Omega$ is the constant angular frequency of the GW, $\boldsymbol{n}$ is a unit vector specifying the direction of propagation of the GW, $H_{ij}$ is the spatial part of the gravitational plane wave amplitude tensor and $\Phi_0$ the phase offset of the GW with respect to some reference time. \newline

Equation \eqref{eq:g_func_trig} contains two simplifying assumptions which must be recognised:  
\begin{enumerate}
	\item The GW is a monochromatic plane wave.
	\item The pulsar locations are constant with respect to the observer 
\end{enumerate}
Regarding point (i), studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves according to \citep[e.g.][]{Sesana2010},
\begin{equation}
	\Delta f_{\rm gw} = 0.05 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^{8.5} M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{50 \mathrm{~nHz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
	\label{eq:f_evolution}
\end{equation}
where $M_{\rm c}$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years. A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$. Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^{8.5} M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 50 \mathrm{~nHz}]^{11/3} (T_{\rm obs}/10 \mathrm{yr})^2 \gtrsim 20$ for $\Delta f_{\rm gw} \gtrsim 1/T_{\rm obs}$. The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}. We are therefore justified in treating the GW source as monochromatic as a first pass in this introductory paper \citep{Sesana10,Sesana2010,Ellis2012ApJ}. \newline
